---
title: "Introduction to R"
author:
  - Bobbur Abhilash Chowdary
  - IIM Calcutta
date: August 2018
output: 
  beamer_presentation:
    keep_tex: true
    theme: CambridgeUS
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, strip.white = TRUE, comment = NA, results = 'hold')
```

## Why R?

>- It's open source!
>- Need not run behind IT support to get the license renewed (SAS users?)
>- It's one of the fastest growing languages along with Python for Machine Learning and AI
>- Easy integration with GitHub and Latex
>- Reproducable research(?)
>- Imagine with one click your code analyzes your data and generates all the required tables & graphs, and finally generates a nice pdf/word document which you can send it to a journal. This can be done in R.

## Installing R

* Download and install R from https://www.r-project.org/

* User interface of just R is very bad. So we need to install R Studio. There are alternative to R-Studio like Emacs etc. You can also try them. 

* Download and install R-Studio from https://www.rstudio.com/. Free version is good enough.

* R-studio runs on top of R. R-studio relies on R to execute all the commands. R-studio cannot function without R. 

* Don't change the order of installation. First R and then R-Studio. Otherwise sometime it can create problems.

* Try this in console and check output to see if RStudio working
```{r test, eval=TRUE}
1+2
```


## Optional - Git -1

Have you ever been frustated seeing files named Final_Version, New_Final_Version ....? Have you ever wondered what is the difference between Version_1 and Version_2?

Then what you need is a version control system. Git is a popular version control system. Luckily R-Studio has very good interface with Git and GitHub. 

Once again Git is open source and free. One can download Git from https://git-scm.com/. 

Git Clients : Git ~ RStudio : R. There several Git clients out there. You may use any of them (Sourcetree, GitUp, GitHub etc). 

Rstudio does most of the basic stuff a Git clients does. Did I forget to say RStudio has good interface with Git?

## Optional - Git -2

Also create GitHub account so you can store your code online. Private repository facility is free if you register with IIM Cal id. GO through the steps after installing Git.

Step 1 - Open Git Bash app and type the following commands

git config --global user.name 'abcd efgh'  
git config --global user.email 'abcdexy@email.iimcal.ac.in'  
git config --global --list

Step 2 - Go to GitHub.com and create a new repository. Choose default options. Open your repository to find readme.md file. Press "clone or download" button on top right corner and copy the url. 

Step 3 - Open Rstudio >File >New Project >Version Control >Git . Paste the above url and choose the folder you want to install the project in. A new folder with repository name will be created.

## Optional - Git -3

Step 4 - Open the readme.md file. Make some changes and save it. In your Git pane in Rstudio (beside Environment pane) select Readme.md and press commit. Type a commit message in new window and then press commit. Then press push (arrow up) to send the commit to GitHub. Open GitHub.com to find the changes reflected in Readme.md file there. 

Use 'diff' in Git pane to see changes you have made since last commit. To revert to previous commits use the addin below the menu section in RStudio. Git is most useful when you are colaborating with your colleagues. You can copy the code from GitHub to your desktop

## RStudio Basics

'Source' pane shows the scripts. For a new script : RStudio >File >New File >R Script

Script has .R extension. It is the place where you will write all your code. Select the code and press 'Run' option on top right corner to execute it. Shortcut: CTRL + ENTER

The 'Console' pane is where you will find the result of executing the code you have written in script. Always read the output in console especially if it is in **RED**

'Environment' pane is where you will find the datasets you have loaded into the memory.

'Files' pane contains the list of files in the current working directory

## Intro to R


* Display current working directory and set current working directory
 
```{r}
getwd()
setwd("G:/My Drive/Thesis/BGs")
```

* Installing Packages and Loading Packages. Need to install a library only once per computer but have to load it everytime you restart. 

```{r echo = TRUE}
install.packages("data.table")
install.packages("tidyverse")
library(data.table)
library(tidyverse)
```
```{r include=FALSE, eval = TRUE}
library(data.table)
library(tidyverse)
```

## Variable names

- R is case sensitive. 'Data' is not same as 'data'.

- Names can have ' . ' (dot), '_', letters and numbers. 

- A good coding convention is to always start with a letter.

- Press 'ALT + SHIFT + K' for list of Keyboard shortcuts

- whenever you find '+' in the console that implies R is waiting for you to complete your command. Try typing `1+` in console.

- Always use '<-' for assigning. '=' also works but sometimes it won't.

- Missing values in R are called 'NA'. The only way to do conditional tests is "is.na". 
```{r}
x =3; y=4; x==y
x =NA; y=NA; x==y
x <- c(1,2,3,4,5,6,NA); x>3; x<4
```

##mtcars data set

mtcars is a default data set in R. We will use this extensively.

mtcars dataset details  
[, 1] mpg  Miles/(US) gallon  
[, 2] cyl  Number of cylinders   
[, 3] disp Displacement (cu.in.)   
[, 4] hp   Gross horsepower  
[, 5] drat Rear axle ratio  
[, 6] wt   Weight (1000 lbs)  
[, 7] qsec 1/4 mile time  
[, 8] vs   Engine (0 = V-shaped, 1 = straight)  
[, 9] am   Transmission (0 = automatic, 1 = manual)   
[,10] gear Number of forward gears  
[,11] carb Number of carburetors  

##Checking out the data set

head() and tail() functions provide the first and last 'n' lines of a given dataset respectively . summary() function summarises and str() gives an idea of structure of data set.
```{r var}
var1 <- head(mtcars,10) 
var2 <- tail(ggplot2::mpg,10)
View(var1) # capital V in View
var1 <- summary(mtcars)
View(var1)
str(mtcars)
```
Common types of variables.  
int - integer;  dbl - real number  
chr - character vector; date - date  
fctr - factor; lgl - logical  


## Column names
Finding Column names, renaming all column names and changing just one column name
```{r}
colnames(var1)
colnames(var1) <- c("mpg", "cyl", "disp", "hp", "drat", 
                "wt", "qsec", "vs", "am", "gear1", "carb")
setnames(var1, "gear1", "gear")
```

## Cleaning Environment and Console 

* Removing a specific data set 
```{r}
rm(var1, var2)
```

* Removing all data sets
```{r}
rm(list = ls())
```

* Clear Console

```{r eval=FALSE, echo=TRUE}
cat("\014")
```


## Reading and writing files
```{r, eval = TRUE}
var0 <- mtcars
write.csv(var0, file = "Output/data.csv")
write.table(var0, file = "Output/data1.txt", sep = "|")
data <- read.csv(file = "Output/data.csv", row.names = 1)
data1 <- read.table("Output/data1.txt", 
                    header = TRUE, sep = "|")
data2 <- read.csv(file = "Output/data.csv")
download.file("https://www.nseindia.com/content/historical/EQUITIES/2018/JUL/cm30JUL2018bhav.csv.zip",
              "NSE BhavCopy/cm30JUL2018bhav.csv.zip")
```

For better import funtions refer 'readr' package. Use 'readxl' package for reading Excel data sheets. 
Use 'haven' & 'foreign' packages for reading SAS, STATA and SPSS files.

## Basic data operations

Most operations in data cleaning process fall into these categories

filter - selecting observations/rows  
arrange - ordering the observations  
select - selecting columns  
mutate - creating new variables from existing ones  
summarize - summary  

The above five are functions in 'dplyr' package. 'dplyr' and a couple of other packages together are called 'tidyverse' which we already loaded.

## dplyr - filter

Use '==' when its a test/question. Use '=' for assigning/informing. We use '==' below because we want to ask "if gear = 4 then select" which is a test.

```{r}
mtcars1 <- filter(mtcars, gear == 4, cyl == 6)
mtcars2 <- filter(mtcars, gear == 4 & cyl == 6) # =mtcars1
mtcars3 <- filter(mtcars, gear == 4 | cyl == 6) #OR operator
mtcars4 <- filter(mtcars, gear == 4 | gear == 6)
mtcars5 <- filter(mtcars, gear %in% c(4,6)) # =mtcars4
```
All 'dplyr operations dont affect the original dataset. mtcars still has 32 observations. Try 'mtcars' on the left side of the above equation. Common operators recognized in R - ">, <, >=, <=, != (not equal) and == "

## dplyr - Arrange and select

Similar to filter(), first input to arrange() is dataset name followed by column names for sorting. use desc() for descing order.
```{r}
arrange(mtcars, mpg, desc(cyl))
```

Selecting columns uses select() function which is similar to arrange() and filter() functions

```{r}
select(mtcars, mpg, cyl)
select(mtcars, -mpg)
select(mtcars, vs, everything()) #rearranging
```

## dplyr - creating variables

Similarly use 'mutate' for creating new variables

```{r}
mutate(mtcars, mpg_per_hp = mpg/hp, mpg_per_wt = mpg/wt, 
       mpg_per_wt1 = mpg_per_hp*hp/wt)
transmutate(mtcars, mpg_per_hp = mpg/hp, mpg_per_wt = mpg/wt, 
       mpg_per_wt1 = mpg_per_hp*hp/wt)
```

You can use the following operators in mutate - '+, -, /, *, ^,  
%% (reminder e.g. - 5 %% 2 =1), %/% (e.g. - 5 %/% 2 = 2), 
sum(), cumsum(), prod() 


## dplyr - summarize() and Pipes

```{r}
by_gear <- group_by(mtcars, gear, carb)
summarise(by_gear, mean_milage = mean(mpg))
```

Pipes are very useful to write good looking code. The most used Pipe is %>%. Pipes also reduce the need to save intermediaries.

x %>% f(y) is equvalent to f(x,y). Above example can be rewritten as...
```{r}
mtcars %>% group_by(gear, carb) %>% 
  summarise(mean_milage = mean(mpg))
```
```{r}
ungroup(mtcars)
```


## ggplot

```{r, eval= TRUE, fig.height= 5 }
ggplot(mtcars ) + geom_point(aes(x= disp, y = mpg))
```

---
```{r, eval= TRUE, message = FALSE, fig.align= 'center', fig.height = 5.5 }
ggplot(mtcars, aes(x= disp, y = mpg))+geom_point(aes(
  size = hp,color = gear)) + geom_smooth()
```

---
```{r, eval= TRUE,  fig.align= 'center' , fig.height = 5.5}
ggplot(mtcars)+ geom_bar(mapping = aes(x = gear))
```

---
```{r, eval= TRUE,  fig.align= 'center', fig.height = 5.5}
ggplot(mtcars)+ geom_histogram(mapping=aes(x=hp),bins=15)
```

## Transforming data (tidyr package) - creating dataset
```{r eval= TRUE}
data <- data.frame(alpha = rep(LETTERS[1:3], 2), 
name = sort(rep(c("bobbur","abhi"),3)), year=2011:2016)
```
```{r echo = FALSE, eval=TRUE}
data
```

## Transforming data - spread()

```{r eval = TRUE}
spread(data, key = name, value = year) 

```

```{r echo = FALSE, eval = TRUE}
data = spread(data, key = name, value = year) 
```

## Transforming data - gather()

```{r eval = TRUE}
gather(data, abhi, bobbur, key = name, value = year)
```
```{r echo = FALSE, eval = TRUE}
data <- gather(data, abhi, bobbur, key = name, value = year)
```

## Others - Joining datasets - rbind

```{r eval=TRUE}
data1 <- data
data1$year <- data1$year + 1000
data2 <- rbind(data, data1)
head(data2,8)
```

## Joining datasets - merge

```{r eval=TRUE}
merge(data,data1, by = c("alpha","name"))
```

## IF ELSE, IFELSE and for loops
```{r eval = TRUE}
x=2
if(x==1){print("hello World") } else if(x==2) 
  {print("hello!")} else {print("!")}
data1$dummy <- ifelse(data1$year>3012, 1, 0)
for (i in 1:nrow(data1)){data1$dummy1[i] <- 
                        data1$dummy[i]+1 }
data1$name <- NULL 
```
```{r echo = FALSE, eval= TRUE}
head(data1,3)
```

## Row Selection 

```{r}
data4 <- data.frame(alpha = rep(LETTERS[1:3], 2), num = 1:6)
elim <- c(TRUE, FALSE, TRUE, TRUE, FALSE, FALSE)
data4[elim,]  
data4[num > 3,]  
data4[data4$num >3,]  
```

```{r echo = FALSE, eval= TRUE}
data4 <- data.frame(alpha = rep(LETTERS[1:3], 2), num = 1:6, beta = letters[c(2,2,2,1,1,1)])
elim <- c(TRUE, FALSE, TRUE, TRUE, FALSE, FALSE)
data4[elim,]

```

## Row selection
```{r  eval= TRUE}
data4[data4$num >3,]
data4[c(1,2,2,2),]
```


## Selection of columns

```{r eval= TRUE}
data4[,c(2,3)] %>% head(n=3)
data4[,c('num', 'alpha')] %>% head(n=3)
```

## duplicates

```{r eval = TRUE}
data4[duplicated(data4$alpha),]
data4[!duplicated(data4$beta),]
```

## order

```{r eval = TRUE}
data4[order(data4$alpha),]  %>% head(n=3)
data4[order(data4$alpha, data4$beta),] %>% head(n=3)
```

---

Thank you!

