---
title: "Introduction to R"
author: Bobbur Abhilash Chowdary
date: August 2018
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Why R?

- It's open source!
- Need not run behind IT support to get the license renewed (SAS users?)
- It's one of the fastest growing languages along with python for Machine Learning and AI
- Reproducable research(?) - This presentation is made in R
- Easy integration with GitHub and Latex

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

